---
title: "EDA_NBA_SportVu"
output: html_document
---

This is a simple walkthrough for getting started with the NBA SportVu movement data. The goal is to show how to parse a game file and perform basic EDA.  

This post is inspired by the work of [Savvast Tjortjoglou](http://savvastjortjoglou.com/nba-play-by-play-movements.html) and [Tanya Cashorali](http://tcbanalytics.com/blog/nba-movement-data-R.html#.VogLV5MrKCQ).

***
For the play, I choose the NBA's [top rated play on December 23rd, 2015](http://www.nba.com/video/channels/top_plays/2015/12/23/20151223-top10.nba/). It is a game between San Antonio and Minnesota and the play occurs with 6 minutes left in the third quarter.  

***
###Download the data
Neil Johnson has taken the time to compile the movement data for NBA games at his [github reposistory](https://github.com/neilmj/BasketballData/tree/master/2016.NBA.Raw.SportVU.Game.Logs).

To get this game, you will need to download the file.

```{r,eval=FALSE}
wget https://github.com/neilmj/BasketballData/blob/master/2016.NBA.Raw.SportVU.Game.Logs/12.23.2015.SAS.at.MIN.7z?raw=true
```

Unzip this file and you should end up with a file named: **0021500431.json**

***
###Reading the data into R

To read this file, first download the _functions.R file in [my github repository](https://github.com/rajshah4/NBA_SportVu) for this project.

```{r}
library(RCurl)
library(jsonlite)
library(dplyr)
library(plotly)
source("_functions.R")
```

The sportvu_convert_json function takes the json file and converts it into a data frame. For this game, the function takes about 3 minutes to convert the file. The resulting data frame is about 2.6 million observations by 13 variables.

```{r}
all.movements <- sportvu_convert_json("data/0021500431.json")
str(all.movements)
```

***
###Finding a specific play
The specific play we are interested in has the event ID #303.
The NBA site has both the [video](http://on.nba.com/1R7pmZF) and [movement data](http://on.nba.com/1YWXb3M) available. The movement data shows the play:  
<iframe height=600 src="http://stats.nba.com/movement/#!/?GameID=0021500431&GameEventID=303" style="border:none;width:100%;height:100% seamless="seamless""></iframe>


***
###Extract movement for one player
The sportvu data provides movement data for every player and the ball. As an example, lets look at the movement of Ginobili for this play.
```{r}
##Extract all data for event ID 303
id303 <- all.movements[which(all.movements$event.id == 303),]
##Extract all data for Ginobili on event ID #303
ginobili <- all.movements[which(all.movements$lastname == "Ginobili" & all.movements$event.id == 303),]
```

This data can be visualized to show how Ginobili moves around the court. The colors represent three different time ranges of movement. The y axis is the length of the court. An NBA court is 94 feet by 50 feet. (Savvast Tjortjoglou takes the time to plot this on a basketball court background image.)
```{r}
p <- plot_ly(data = ginobili, x = x_loc, y = y_loc, mode = "markers", color=cut(ginobili$game_clock, breaks=3)) %>% 
    layout(xaxis = list(range = c(0, 100)), 
           yaxis = list(range = c(0, 50))) 
p
```

***
###Get distance travelled for one player
I have a simple function to get the distance a player travels:
```{r}
travelDist(ginobili$x_loc, ginobili$y_loc)
```

***
###Get speed of a player
Building off the distance, it is possible to calculate the speed of a player.

```{r}
seconds = max(ginobili$game_clock) - min(ginobili$game_clock)
speed = travelDist(ginobili$x_loc, ginobili$y_loc)/seconds  #in feet per second
speed
```

***
###Get distance for all the players for a specific event 
The next step is generalizing this approach to all the players.
```{r}
player.groups <- group_by(id303, lastname)
dist.traveled.players <- summarise(player.groups, totalDist=travelDist(x_loc, y_loc),playerid = max(player_id))
arrange(dist.traveled.players, desc(totalDist))
```

***
###Get distance for all the players for the entire game
This part extends the measurement to an entire game for all the players. For this game, the most active players went a little over 2 miles, which makes sense.
```{r}
deduped.data <- unique( all.movements[ , 1:12 ] )  ##This takes about 30 seconds to run
player.groups <- group_by(deduped.data, lastname)
dist.traveled.players <- summarise(player.groups, totalDist=travelDist(x_loc,y_loc),playerid = max(player_id))
total <- arrange(dist.traveled.players, desc(totalDist))
total
```

***
###Get the distance between a player and the ball for one event
A more interesting use of the data is to see how distances between people and the ball change over time. This code shows you how to get the distance between two parties for an event. The example here uses Ginobili and the ball.

```{r}
ginobili <- all.movements[which((all.movements$lastname == "Ginobili"| all.movements$lastname == "ball") & all.movements$event.id == 303),]
#Get distance for each player/ball
distgino <- ginobili %>% filter (lastname=="Ginobili") %>% select (x_loc,y_loc) 
distball <- ginobili %>% filter (lastname=="ball") %>% select (x_loc,y_loc) 
distlength <- 1:nrow(distgino)
#Use the R function dist for calculating distance
distsdf <- unlist(lapply(distlength,function(x) {dist(rbind(distgino[x,], distball[x,]))}))
#Add the game_clock
ball_distance <- ginobili %>% filter (lastname=="ball") %>% select (game_clock) %>% mutate(distance=distsdf)
plot_ly(data = ball_distance, x=game_clock, y=distsdf,mode = "markers")
```

***
###Get the distance between a player and the ball for one event (using functions)
This part uses the same logic as above, but with functions I created to make it cleaner.
```{r}
#Get Clock Info
clockinfo <- get_game_clock("Ginobili",303)
#Get Distance
playerdistance <- player_dist("Ginobili","ball",303)
#Plot
plot_ly(data = clockinfo, x=game_clock, y=playerdistance,mode = "markers")
```

***
### Get the distance between all players and the ball for one event
This section generalizes the code to view the distance between all the players and the ball. The plot can be a bit messy (and does not show in Rpubs), but its an interesting way to see the interactions between players and the ball.
```{r}
pickplayer <- "ball"
pickeventID <- 303

#Get all the players
players <- all.movements %>% filter(event.id==pickeventID) %>% select(lastname) %>% distinct(lastname)
#Calculate distance
bigdistance <- lapply(list(players$lastname)[[1]],function (x){player_dist(pickplayer,x,pickeventID)})
bigdistancedf <- as.data.frame(do.call('cbind',bigdistance))
colnames(bigdistancedf) <- list(players$lastname)[[1]]
#Get Clock Info
clockinfo <- get_game_clock(pickplayer,pickeventID)
bigdistancedf$game_clock <- clockinfo$game_clock
head(bigdistancedf)

##Plot with plotly - not elegant but shows you one way to visualize the data
for(i in 1:(ncol(bigdistancedf)-1)){
if(i==1){
  pString<-"p <- plot_ly(data = bigdistancedf, x = game_clock, y = bigdistancedf[,1], name = colnames(bigdistancedf[1]))"
} else {
  pString<-paste(pString, " %>% add_trace(y =",  eval(paste("bigdistancedf[,",i,"]",sep="")),", name=", eval(paste("colnames(bigdistancedf[", i,"])",sep="")), ")", sep="")
}
}
eval(parse(text=pString))
print(p)
```

***
###Get a distance matrix between all the players for one eventID
The movement data also allows for the analysis of the distance between players. For example, if you are interested in the distance between LaVine and Ginobili for a certain play. This function calculates a matrix of all the distances between players and the ball.

```{r}
pickeventID <- 303
players_matrix <- player_dist_matrix(pickeventID)
str(players_matrix)
```

***
###Credits
I hope this has been a useful introduction to working with the NBA movement data. For more of my explorations on the NBA data you can see my [NBA Github repo](https://github.com/rajshah4/NBA_SportVu). You can find more information about me, [RajivShah](http://www.rajivshah.com) or my other [projects](http://projects.rajivshah.com) or find me on [Twitter](http://twitter.com/rajcs4).
